### 2018-04-04 paper.Rmd
updates to outline during call

abstract: |
  The Child Language Data Exchange System (CHILDES) has played a critical role in research on child language development. However, traditional methods of inquiry into the data have often relied on a command line interface to extract developmental indices or custom parsers to extract data for computational modeling. Now, advances in open-source software allow for novel approaches in working with this dataset. We introduce childes-db, a tool that improves data accessibility and usability by offering novel interfaces to CHILDES, including browsable web applications and an R application programming interface.  Along with a versioned infrastructure that facilitates reproduction of past analyses, these interfaces lower barriers to analyzing naturalistic parent-child language, allowing for a wider range of researchers in language and cognitive development to easily leverage CHILDES in their work.

@@

Yet there are some challenges working with CHILDES, both for students and for advanced users. For beginners with a lower level of technical experience, mastering the system can be tricky. The CHILDES ecosystem uses a special file format (CHAT), which is stored in a plain text format but includes structured annotations grouped into tiers (stored on separate lines) so that information about utterances can be stored with accompanying information such as the phonological or morphological structure of the utterance. These files are analyzed using a command-line tool (CLAN) with a host of specific arguments that allow users to count word frequencies, compute statistics (e.g., mean length of utterance, or MLU). While this system is flexible and powerful, mastering the CHAT codes and the CLAN tool can be daunting to navigate for a beginner or student. These technical barriers decrease the ease of casual exploration by a novice researcher or in a classroom exercise.

@@

# The Architecture of childes-db

We first present the design and technical approach of childes-db, then describe the interactive analysis tools provided by our web-based frontend.

## Design and technical approach

@@

Other tables in the database comprise a hierarchical system of collections built on top of tokens -- *utterance*, *transcription*, *corpus*, and *collection* -- which stores attributes appropriate for each level of description. Every entity includes attributes that link it to all higher-order collections, e.g., an utterance lists the transcript, corpus, and collection to which it belongs.  An *utterance* contains one or more words and includes fields such as the utterance type (declarative, question, etc.), total number of tokens, and the total number one morphemes. [[ How are we handling the long tail of tiers?]]. A *transcription* consists of one or more utterances and includes the date collected, the name of the target child and age in days (if defined), and the filename from CHILDES. A *corpus* consists of one or more transcripts, corresponding to well-known collections like the Brown or Providence corpus. Finally, a *collection* is a superordinate collection of corpora generally corresponding to a geographic region, following the convention in CHILDES. Because every entity can be linked to a top-level collection, childes-db places content from all languages in the same table for a given level of abstraction.

@@

### Technical Implementation

The above tabular datastore is implemented as a MySQL database.  As an industry-standard, open-source relational database server, the above database can be accessed directly from a wide range of programming languages. The childes-db project provides hosted read-only databases for direct access and for `childesr` (described below) as well as compressed .sql exports for local installation. While the former is appropriate for casual users, local installation can provide performance gains (allowing a user to access the database on their local machine or on their network) as well as write support in case users wish to store derived information in the same database.

[[Details of extraction]]
[[Details of testing: Why trust our parse over your in-house library?]]

@@

@@

@@

@@

### Population Viewer
The population viewer is a tool that helps users understand the relative sizes and temporal extents of various corpora by displaying the total number of utterances or tokens in the available corpora at particular ages. In many instances the size of a given corpus may inform researchers’ decisions as to what is the best target dataset for analysis. <!-- note that the graphs don't currently include a grouping factor, so adding multiple corpora generates one line, so you can't actually compare two corpora in the same visualization -->

# The `childesr` Package

@@

## Package Details

@@

## Worked Examples

@@

#### Using shiny

@@

# Teaching with childes-db

## In class demonstrations

@@

## Tutorials and assignments

@@

reiterate importance

# Acknowledgements
- Brian MacWhinney for CHILDES
- Melissa Kline for ClanToR

\newpage

>>>>

abstract: |
  The Child Language Data Exchange System (CHILDES) has played a critical role in research on child language development. However, traditional methods of inquiry into the data have often relied on a command line interface to extract developmental indices or custom parsers to extract data for computational modeling. Now, advances in open-source software allow for novel approaches in working with this dataset. We introduce childes-db, a database-formatted mirror of CHILDES that improves data accessibility and usability by offering novel interfaces, including browsable web applications and an R application programming interface (API).  Along with a versioned infrastructure that facilitates reproduction of past analyses, these interfaces lower barriers to analyzing naturalistic parent-child language, allowing for a wider range of researchers in language and cognitive development to easily leverage CHILDES in their work.

@@

Yet there are some challenges working with CHILDES, both for students and for advanced users. The CHILDES ecosystem uses a special file format (CHAT), which is stored in a plain text format but includes structured annotations grouped into tiers (stored on separate lines) so that information about utterances can be stored with accompanying information such as the phonological or morphological structure of the utterance. These files are analyzed using a command-line tool (CLAN) with a host of specific arguments that allow users to count word frequencies, compute statistics (e.g., mean length of utterance, or MLU). While this system is flexible and powerful, mastering the CHAT codes and the CLAN tool can be daunting to navigate for a beginner or student. These technical barriers decrease the ease of casual exploration by a novice researcher or in a classroom exercise.

@@

# Design and technical approach

<!-- Remind readers why they're reading this: understanding the structure helps you use and understand the tools -->

<!-- We first present the design and technical approach of childes-db, then describe the interactive analysis tools provided by our web-based frontend.  -->

## Database format

@@

Other tables in the database comprise a hierarchical system of collections built on top of tokens -- *utterance*, *transcription*, *corpus*, and *collection* -- which stores attributes appropriate for each level of description. Every entity includes attributes that link it to all higher-order collections, e.g., an utterance lists the transcript, corpus, and collection to which it belongs.  An *utterance* contains one or more words and includes fields such as the utterance type (declarative, question, etc.), total number of tokens, and the total number one morphemes.

A *transcription* consists of one or more utterances and includes the date collected, the name of the target child and age in days (if defined), and the filename from CHILDES. A *corpus* consists of one or more transcripts, corresponding to well-known collections like the Brown or Providence corpus. Finally, a *collection* is a superordinate collection of corpora generally corresponding to a geographic region, following the convention in CHILDES. Because every entity can be linked to a top-level collection, childes-db places content from all languages in the same table for a given level of abstraction.

@@

## Technical implementation

<!-- [[mention using XML childes representation]] -->

The above tabular datastore is implemented as a MySQL database.  As an industry-standard, open-source relational database server, the above database can be accessed directly from a wide range of programming languages. The childes-db project provides hosted read-only databases for direct access and for `childesr` (described below) as well as compressed .sql exports for local installation. While the former is appropriate for most users, local installation can provide performance gains (allowing a user to access the database on their local machine or on their network) as well as allowing users to store derived information in the same database.

<!-- [[Details of extraction]] -->
<!-- [[Details of testing: Why trust our parse over your in-house library? CLAN TESTING DETAILS - also mention versioning as a reason to trust]] -->

@@

<!-- Continue to make the case that it's important to version! childes changes without warning, corpora get added, etc. -->

@@

### Current annotation coverage

<!-- [[ How are we handling the long tail of tiers? first priority is lexical statistics & morphology, but there's other stuff too]].  -->

<!-- no media links currently, no timing information -->

<!-- this stuff could be added at some point without changing format -->

# Interfaces for Accessing childes-db

We first discuss the childes-db web apps and then introduce the `childesr` package.

@@

<!-- need a good shorthand for the interactive analysis tools: web apps -->

@@

### Population Viewer

The population viewer is a tool that helps users understand the relative sizes and temporal extents of various corpora by displaying the total number of utterances or tokens in the available corpora at particular ages. In many instances the size of a given corpus may inform researchers’ decisions as to what is the best target dataset for analysis. <!-- note that the graphs don't currently include a grouping factor, so adding multiple corpora generates one line, so you can't actually compare two corpora in the same visualization -->

## The `childesr` Package

@@

@@

# Using childes-db: Worked Examples

In this section we give a number of examples of how childes-db can be used in both research and teaching, using both the web apps and the R API.

## Research applications

@@

#### Using web apps

@@

## Teaching with childes-db

### In class demonstrations

@@

### Tutorials and assignments

@@

<!-- reiterate importance -->

<!-- adding more tiers is a goal for the future -->

# Acknowledgements

Thanks to Brian MacWhinney for advice and guidance, and to Melissa Kline for her work on ClanToR, which formed a starting point for our work. This work supported by a Jacobs Advanced Research Fellowship to MCF.
